---
title: "An Analysis of S&P 500 and Commodity Price Correlations"
author: "Reginal Supoint"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Introduction

The purpose of this analysis is to explore the correlation between the S&P 500 index and various commodities such as oil, gold, silver, platinum and palladium. Understanding these correlations can provide insights into market dynamics and investment strategies. 

This dataset includes daily closing prices for the S&P 500, NASDAQ, and several commodities. This analysis will include correlation analysis, time series forecasting and scenario analysis to understand potential future trends.


```{r, include = FALSE, message = FALSE, warning= FALSE}
# Load the packages
library(tidyverse)
library(corrplot)
library(forecast)
library(ggplot2)
library(rugarch)
options(repos = c(CRAN = "https://cloud.r-project.org"))




```


```{R Load-Data, include = FALSE}
data <- read.csv('financial_regression.csv')
head(data)
summary(data)
sum(is.na(data))
data_clean <- na.omit(data)
```

```{r}
# Select relevant columns for correlation analysis
cor_data <- data_clean %>% select(sp500.close, nasdaq.close, oil.close, gold.close, silver.close)

# Calculate the correlation matrix
cor_matrix <- cor(cor_data)

# Print the correlation matrix
print(cor_matrix)

# Visualize the correlation matrix
corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black", tl.srt = 45)
```
``
This correlation map provides a solid snapshot of how different assets are related to each other. We are particularly looking at how the commodities are related to the S&P 500.

Correlations Scores range from -1 to 1, With 1 being the strongest correlation and -1 being the weakest correlation.

## S&P 500 and NASDAQ 
With a score of 0.99 we can see that the S&P 500 and NASDAQ are almost perfectly positively correlated. Therefore, when the S&P 500 moves in a certain direction the NASDAQ is likely to move in the same direction. 

## Correlation Summary:
High Correlation: S&P 500 and NASDAQ, NASDAQ and Gold, S&P 500 and GOLD 

Moderate Correlation: S&P 500 and Oil, NASDAQ and Oil, Oil and Gold, Oil and Silver.

Weak or No Correlation: S&P 500 and Silver, NASDAQ and Silver. 
Due to the Market Dynamics Silver does not move in tandem with the Stock Market. Silver is more influenced by supply and demand dynamics specific to industrial and investment usages versus the stock market. Silver also tend to be highly volatile meaning weak correlation with the stock market. 

```{r}
# Convert the date column to Date type
data_clean$date <- as.Date(data_clean$date, format = "%m/%d/%Y")

# Arrange data by date
data_clean <- data_clean %>% arrange(date)

horizon <- 5 *252

# Fit an ARIMA model to the oil closing prices
model <- auto.arima(data_clean$oil.close, seasonal = FALSE)

# Print the model summary
summary(model)
```

```{r}
# Forecast future values
forecast_values <- forecast(model, h = horizon)

# Create a data frame with the forecast values and dates
forecast_df <- data.frame(
  date = seq.Date(from = max(data_clean$date) + 1, by = "day", length.out = horizon),
  forecast = as.numeric(forecast_values$mean)
)

# Combine the actual and forecast data for plotting
plot_data <- data_clean %>%
  select(date, oil.close) %>%
  rename(actual = oil.close) %>%
  bind_rows(forecast_df %>% rename(actual = forecast))

# Plot the actual and forecasted values
ggplot(plot_data, aes(x = date)) +
  geom_line(aes(y = actual, color = "Actual")) +
  geom_line(data = forecast_df, aes(y = forecast, color = "Forecast")) +
  labs(title = "Oil Price Forecast", x = "Date", y = "Price") +
  scale_color_manual(values = c("Actual" = "blue", "Forecast" = "red"))

```
In order to forecast the movement of stocks for the future I thought the ARIMA model would be the appropriate method. The model uses previous values of the time series in order to predict future outcomes. Which I thought would be an appropriate way to predict the value of Oil in a vacuum; before I go ahead and graph the future values of the other commodities. 

The graph itself is forecasting for the price of oil to drop in 2025. 






```{r}
forecast_commodity <- function(data, column_name, horizon) {
  # Fit ARIMA model
  model <- auto.arima(data[[column_name]], seasonal = FALSE)
  # Forecast future values
  forecast_values <- forecast(model, h = horizon)
  # Create a data frame with the forecast values and dates
  forecast_df <- data.frame(
    date = seq.Date(from = max(data$date) + 1, by = "day", length.out = horizon),
    forecast = as.numeric(forecast_values$mean)
  )
  return(forecast_df)
}

# Forecast for each commodity
oil_forecast <- forecast_commodity(data_clean, "oil.close", horizon)
gold_forecast <- forecast_commodity(data_clean, "gold.close", horizon)
silver_forecast <- forecast_commodity(data_clean, "silver.close", horizon)
platinum_forecast <- forecast_commodity(data_clean, "platinum.close", horizon)
palladium_forecast <- forecast_commodity(data_clean, "palladium.close", horizon)



```

For the models I noticed that all of the prices for the Commodities stayed exactly on a straight line for the prediction that was trying to be made. Which made me consider if I was using the correct model to forecast the price.



```{r}
# Combine forecast data with historical data
combine_data <- function(data, forecast_df, column_name) {
  combined <- data %>%
    select(date, !!sym(column_name)) %>%
    rename(actual = !!sym(column_name)) %>%
    bind_rows(forecast_df %>% rename(actual = forecast))
  return(combined)
}

# Combine data for each commodity
oil_combined <- combine_data(data_clean, oil_forecast, "oil.close")
gold_combined <- combine_data(data_clean, gold_forecast, "gold.close")
silver_combined <- combine_data(data_clean, silver_forecast, "silver.close")
platinum_combined <- combine_data(data_clean, platinum_forecast, "platinum.close")
palladium_combined <- combine_data(data_clean, palladium_forecast, "palladium.close")


```

```{r}
# Plot the actual and forecasted values for all commodities and S&P 500
ggplot() +
  geom_line(data = oil_combined, aes(x = date, y = actual, color = "Oil Actual")) +
  geom_line(data = oil_forecast, aes(x = date, y = forecast, color = "Oil Forecast")) +
  geom_line(data = gold_combined, aes(x = date, y = actual, color = "Gold Actual")) +
  geom_line(data = gold_forecast, aes(x = date, y = forecast, color = "Gold Forecast")) +
  geom_line(data = silver_combined, aes(x = date, y = actual, color = "Silver Actual")) +
  geom_line(data = silver_forecast, aes(x = date, y = forecast, color = "Silver Forecast")) +
  geom_line(data = platinum_combined, aes(x = date, y = actual, color = "Platinum Actual")) +
  geom_line(data = platinum_forecast, aes(x = date, y = forecast, color = "Platinum Forecast")) +
  geom_line(data = palladium_combined, aes(x = date, y = actual, color = "Palladium Actual")) +
  geom_line(data = palladium_forecast, aes(x = date, y = forecast, color = "Palladium Forecast")) +
  labs(title = "Commodity Prices Forecast for the Next 5 Years",
       x = "Date", y = "Price") +
  scale_color_manual(values = c(
    "Oil Actual" = "blue", "Oil Forecast" = "darkblue",
    "Gold Actual" = "red", "Gold Forecast" = "darkred",
    "Silver Actual" = "green", "Silver Forecast" = "darkgreen",
    "Platinum Actual" = "purple", "Platinum Forecast" = "violet",
    "Palladium Actual" = "orange", "Palladium Forecast" = "darkorange"
  )) +
  theme_minimal()

```


```{r, echo=FALSE}
# Function to adjust ARIMA models based on scenarios
scenario_forecast <- function(data, column_name, horizon, scenario) {
  if (scenario == "boom") {
    # Increase the level by 10% to simulate economic boom
    adjustment <- 1.10
  } else if (scenario == "recession") {
    # Decrease the level by 10% to simulate economic recession
    adjustment <- 0.90
  } else {
    adjustment <- 1.00
  }

  model <- auto.arima(data[[column_name]], seasonal = FALSE)
  forecast_values <- forecast(model, h = horizon)
  adjusted_forecast <- forecast_values$mean * adjustment

  forecast_df <- data.frame(
    date = seq.Date(from = max(data$date) + 1, by = "day", length.out = horizon),
    forecast = as.numeric(adjusted_forecast)
  )
  return(forecast_df)
}

# Set the forecast horizon to 5 years
horizon <- 5 * 252

# Generate scenario-based forecasts for each commodity
oil_forecast_boom <- scenario_forecast(data_clean, "oil.close", horizon, "boom")
oil_forecast_recession <- scenario_forecast(data_clean, "oil.close", horizon, "recession")
gold_forecast_boom <- scenario_forecast(data_clean, "gold.close", horizon, "boom")
gold_forecast_recession <- scenario_forecast(data_clean, "gold.close", horizon, "recession")
silver_forecast_boom <- scenario_forecast(data_clean, "silver.close", horizon, "boom")
silver_forecast_recession <- scenario_forecast(data_clean, "silver.close", horizon, "recession")
platinum_forecast_boom <- scenario_forecast(data_clean, "platinum.close", horizon, "boom")
platinum_forecast_recession <- scenario_forecast(data_clean, "platinum.close", horizon, "recession")
palladium_forecast_boom <- scenario_forecast(data_clean, "palladium.close", horizon, "boom")
palladium_forecast_recession <- scenario_forecast(data_clean, "palladium.close", horizon, "recession")





# Combine scenario-based forecast data with historical data
combine_scenario_data <- function(data, baseline_forecast, boom_forecast, recession_forecast, column_name) {
  combined <- data %>%
    select(date, !!sym(column_name)) %>%
    rename(actual = !!sym(column_name)) %>%
    bind_rows(baseline_forecast %>% rename(baseline = forecast)) %>%
    bind_rows(boom_forecast %>% rename(boom = forecast)) %>%
    bind_rows(recession_forecast %>% rename(recession = forecast))
  return(combined)
}

# Combine data for each commodity under both scenarios
oil_combined <- combine_scenario_data(data_clean, oil_forecast, oil_forecast_boom, oil_forecast_recession, "oil.close")
gold_combined <- combine_scenario_data(data_clean, gold_forecast, gold_forecast_boom, gold_forecast_recession, "gold.close")
silver_combined <- combine_scenario_data(data_clean, silver_forecast, silver_forecast_boom, silver_forecast_recession, "silver.close")
platinum_combined <- combine_scenario_data(data_clean, platinum_forecast, platinum_forecast_boom, platinum_forecast_recession, "platinum.close")
palladium_combined <- combine_scenario_data(data_clean, palladium_forecast, palladium_forecast_boom, palladium_forecast_recession, "palladium.close")




# Plot the scenario-based forecasts for all commodities
ggplot() +
  geom_line(data = oil_combined, aes(x = date, y = actual, color = "Oil Actual")) +
  geom_line(data = oil_forecast, aes(x = date, y = forecast, color = "Oil Baseline")) +
  geom_line(data = oil_forecast_boom, aes(x = date, y = forecast, color = "Oil Boom")) +
  geom_line(data = oil_forecast_recession, aes(x = date, y = forecast, color = "Oil Recession")) +
  geom_line(data = gold_combined, aes(x = date, y = actual, color = "Gold Actual")) +
  geom_line(data = gold_forecast, aes(x = date, y = forecast, color = "Gold Baseline")) +
  geom_line(data = gold_forecast_boom, aes(x = date, y = forecast, color = "Gold Boom")) +
  geom_line(data = gold_forecast_recession, aes(x = date, y = forecast, color = "Gold Recession")) +
  geom_line(data = silver_combined, aes(x = date, y = actual, color = "Silver Actual")) +
  geom_line(data = silver_forecast, aes(x = date, y = forecast, color = "Silver Baseline")) +
  geom_line(data = silver_forecast_boom, aes(x = date, y = forecast, color = "Silver Boom")) +
  geom_line(data = silver_forecast_recession, aes(x = date, y = forecast, color = "Silver Recession")) +
  geom_line(data = platinum_combined, aes(x = date, y = actual, color = "Platinum Actual")) +
  geom_line(data = platinum_forecast, aes(x = date, y = forecast, color = "Platinum Baseline")) +
  geom_line(data = platinum_forecast_boom, aes(x = date, y = forecast, color = "Platinum Boom")) +
  geom_line(data = platinum_forecast_recession, aes(x = date, y = forecast, color = "Platinum Recession")) +
  geom_line(data = palladium_combined, aes(x = date, y = actual, color = "Palladium Actual")) +
  geom_line(data = palladium_forecast, aes(x = date, y = forecast, color = "Palladium Baseline")) +
  geom_line(data = palladium_forecast_boom, aes(x = date, y = forecast, color = "Palladium Boom")) +
  geom_line(data = palladium_forecast_recession, aes(x = date, y = forecast, color = "Palladium Recession")) +
  labs(title = "Commodity Prices Scenario-Based Forecast for the Next 5 Years",
       x = "Date", y = "Price") +
  scale_color_manual(values = c(
    "Oil Actual" = "blue", "Oil Baseline" = "darkblue", "Oil Boom" = "lightblue", "Oil Recession" = "blue4",
    "Gold Actual" = "red", "Gold Baseline" = "darkred", "Gold Boom" = "lightcoral", "Gold Recession" = "red4",
    "Silver Actual" = "green", "Silver Baseline" = "darkgreen", "Silver Boom" = "lightgreen", "Silver Recession" = "green4",
    "Platinum Actual" = "purple", "Platinum Baseline" = "darkviolet", "Platinum Boom" = "mediumorchid", "Platinum Recession" = "orchid4",
    "Palladium Actual" = "orange", "Palladium Baseline" = "darkorange", "Palladium Boom" = "gold", "Palladium Recession" = "orange4"
  )) +
  theme_minimal()

```
To Test if my model was working the way I wanted it to I decided to graph the prices of the Commodities based on two scenarios. One which predicted a 10% increase in all the prices represented as a "boom" and another that represented a 10% decrease represented as a "recession". This confirmed to me the the graphs were indeed working, but I decided to test out a different statistical model  instead. 

```{r, echo=FALSE}


# Install and load necessary packages
install.packages("rugarch")
install.packages("xts")
library(rugarch)
library(xts)
library(ggplot2)
library(dplyr)
library(tibble)

# Ensure data is clean and without NAs or Infs
oil_returns <- data_clean$oil.close  # Extract oil close prices
oil_returns <- na.omit(oil_returns)  # Remove NA values
oil_returns <- oil_returns[!is.infinite(oil_returns)]  # Remove Inf values
oil_returns <- oil_returns[!is.nan(oil_returns)]  # Remove NaN values

# Convert oil_returns to a time series object
oil_returns <- xts(oil_returns, order.by = data_clean$date[!is.na(oil_returns)])

# Specify the GARCH model
spec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0))
)

# Fit the GARCH model to your data
oil_garch_fit <- ugarchfit(spec = spec, data = oil_returns)

# Forecast volatility using the fitted GARCH model
oil_garch_forecast <- ugarchforecast(oil_garch_fit, n.ahead = 30)

# Extract the forecasted volatility (sigma)
forecast_volatility <- sigma(oil_garch_forecast)

# Prepare the forecast data
forecast_df <- data.frame(
  Date = as.Date(index(forecast_volatility)),  # Ensure oil_garch_forecast has date index
  Volatility = as.numeric(forecast_volatility)  # Ensure sigma(oil_garch_forecast) returns numeric values
)

# Add confidence intervals for volatility
forecast_df <- forecast_df %>%
  mutate(
    Lower_CI = Volatility * 0.95,
    Upper_CI = Volatility * 1.05
  )

# Historical data for context
historical_volatility <- data.frame(
  Date = as.Date(index(oil_garch_fit@fit$sigma)),  # Ensure oil_garch_fit has a date index
  Volatility = as.numeric(oil_garch_fit@fit$sigma)  # Ensure sigma(oil_garch_fit) returns numeric values
)

# Add a column to distinguish historical data
historical_volatility <- historical_volatility %>%
  mutate(Type = "Historical")

# Add a column to distinguish forecast data
forecast_df <- forecast_df %>%
  mutate(Type = "Forecast")

# Combine historical and forecasted volatility
combined_data <- bind_rows(
  historical_volatility %>%
    filter(Date >= as.Date("2024-01-01")),  # Focus on recent historical data
  forecast_df
)

# Plot the combined data with confidence intervals
ggplot(combined_data, aes(x = Date, y = Volatility, color = Type)) +
  geom_line(size = 1.2) +
  geom_ribbon(
    data = forecast_df,
    aes(x = Date, ymin = Lower_CI, ymax = Upper_CI, fill = Type),
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  labs(
    title = "Oil GARCH Volatility Forecast with Historical Context",
    x = "Date",
    y = "Volatility",
    color = "Type",
    fill = "Type"
  ) +
  scale_color_manual(values = c("Historical" = "black", "Forecast" = "blue")) +
  scale_fill_manual(values = c("Forecast" = "blue")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 14)
  )

```

Another model I used is called the GARCH model, which is used to estimate the volatility of financial returns. Typically, it is used in markets to model and forecast the volatility of different assets. 
Looking at the trend line it is clear that volatility is expected to steady increase over time. You can also see that the interval increase over time, generally as the interval becomes wider it represents greater uncertainty in the forecast. Therefore, this sugests that there is uncertainty in the future. This could represent greater risks in the market for traders and investors in oil related assets.



```{r, echo=FALSE}
# Ensure the data is clean and without NAs or Infs for each commodity
prepare_data <- function(data) {
  data <- na.omit(data)  # Remove NA values
  data <- data[!is.infinite(data)]  # Remove Inf values
  data <- data[!is.nan(data)]  # Remove NaN values
  return(data)
}

# Prepare data for all commodities
gold_returns <- prepare_data(data_clean$gold.close)
silver_returns <- prepare_data(data_clean$silver.close)
platinum_returns <- prepare_data(data_clean$platinum.close)
palladium_returns <- prepare_data(data_clean$palladium.close)

# Convert to time series objects
gold_returns <- xts(gold_returns, order.by = data_clean$date[!is.na(gold_returns)])
silver_returns <- xts(silver_returns, order.by = data_clean$date[!is.na(silver_returns)])
platinum_returns <- xts(platinum_returns, order.by = data_clean$date[!is.na(platinum_returns)])
palladium_returns <- xts(palladium_returns, order.by = data_clean$date[!is.na(palladium_returns)])

# Specify the GARCH model
spec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model = list(armaOrder = c(0, 0))
)

# Fit the GARCH models
gold_garch_fit <- ugarchfit(spec = spec, data = gold_returns)
silver_garch_fit <- ugarchfit(spec = spec, data = silver_returns)
platinum_garch_fit <- ugarchfit(spec = spec, data = platinum_returns)
palladium_garch_fit <- ugarchfit(spec = spec, data = palladium_returns)

# Forecast volatilities
gold_garch_forecast <- ugarchforecast(gold_garch_fit, n.ahead = 30)
silver_garch_forecast <- ugarchforecast(silver_garch_fit, n.ahead = 30)
platinum_garch_forecast <- ugarchforecast(platinum_garch_fit, n.ahead = 30)
palladium_garch_forecast <- ugarchforecast(palladium_garch_fit, n.ahead = 30)

# Extract forecasted volatilities
gold_forecast_vol <- sigma(gold_garch_forecast)
silver_forecast_vol <- sigma(silver_garch_forecast)
platinum_forecast_vol <- sigma(platinum_garch_forecast)
palladium_forecast_vol <- sigma(palladium_garch_forecast)

```

```{r, echo=FALSE}
# Prepare forecast data for all commodities
prepare_forecast_data <- function(forecast_volatility, type_label) {
  data.frame(
    Date = as.Date(index(forecast_volatility)),
    Volatility = as.numeric(forecast_volatility),
    Type = type_label
  ) %>%
  mutate(
    Lower_CI = Volatility * 0.95,
    Upper_CI = Volatility * 1.05
  )
}

gold_forecast_df <- prepare_forecast_data(gold_forecast_vol, "Gold Forecast")
silver_forecast_df <- prepare_forecast_data(silver_forecast_vol, "Silver Forecast")
platinum_forecast_df <- prepare_forecast_data(platinum_forecast_vol, "Platinum Forecast")
palladium_forecast_df <- prepare_forecast_data(palladium_forecast_vol, "Palladium Forecast")

# Historical volatilities for context
gold_historical_vol <- data.frame(
  Date = as.Date(index(gold_garch_fit@fit$sigma)),
  Volatility = as.numeric(gold_garch_fit@fit$sigma),
  Type = "Gold Historical"
)

silver_historical_vol <- data.frame(
  Date = as.Date(index(silver_garch_fit@fit$sigma)),
  Volatility = as.numeric(silver_garch_fit@fit$sigma),
  Type = "Silver Historical"
)

platinum_historical_vol <- data.frame(
  Date = as.Date(index(platinum_garch_fit@fit$sigma)),
  Volatility = as.numeric(platinum_garch_fit@fit$sigma),
  Type = "Platinum Historical"
)

palladium_historical_vol <- data.frame(
  Date = as.Date(index(palladium_garch_fit@fit$sigma)),
  Volatility = as.numeric(palladium_garch_fit@fit$sigma),
  Type = "Palladium Historical"
)

# Combine all data
combined_data <- bind_rows(
  gold_historical_vol, gold_forecast_df,
  silver_historical_vol, silver_forecast_df,
  platinum_historical_vol, platinum_forecast_df,
  palladium_historical_vol, palladium_forecast_df
)

```

```{r}
# Plot the combined data with confidence intervals
ggplot(combined_data, aes(x = Date, y = Volatility, color = Type)) +
  geom_line(size = 1.2) +
  geom_ribbon(
    data = combined_data %>% filter(str_detect(Type, "Forecast")),
    aes(x = Date, ymin = Lower_CI, ymax = Upper_CI, fill = Type),
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  labs(
    title = "Commodity Volatility Forecasts with Historical Context",
    x = "Date",
    y = "Volatility",
    color = "Type",
    fill = "Type"
  ) +
  scale_color_manual(values = c(
    "Gold Historical" = "red", "Gold Forecast" = "darkred",
    "Silver Historical" = "blue", "Silver Forecast" = "darkblue",
    "Platinum Historical" = "green", "Platinum Forecast" = "darkgreen",
    "Palladium Historical" = "orange", "Palladium Forecast" = "darkorange"
  )) +
  scale_fill_manual(values = c(
    "Gold Forecast" = "darkred",
    "Silver Forecast" = "darkblue",
    "Platinum Forecast" = "darkgreen",
    "Palladium Forecast" = "darkorange"
  )) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 14)
  )

```
## Conclusion: 
Gold shows a relatively stable potential forecast compared to its historical trends, volatility remains steady.Palladium shows some extreme levels of fluctuations, but the forecast remains high and stable. Platinum displays historical volatility but the forecast suggest stability. Silver showed that both historical and forecasted volatility are minimal compared to the other commodities.   

## Important Observations 
Palladium's historical volatility exhibits sharp spikes that are likely influenced by disruptions in the market or other factors. The forecast predicts this volatility to stabilize but stay around the 100 - 150 range. 

Gold has the most stable historical and forecasted volatility, reflecting its reputation as a safe investment. 

Platinum demonstrated moderate historical fluctuations, with its forecast suggesting a reduction in volatility in the future. 

Silver has minimal volatility, with historical and forecasted levels being closely aligned, which could imply silver has a stable market.

## Potential Insights

As expected, gold and silver appear to be less risky investments due to their stable volatility trends, make them safer options for risk-averse investors. 

Platinum shows moderate volatility offering a medium-risk option for investors

Palladium's volatility signals opportunities for high-risk, high-reward strategies but will often require careful consideration of the market.
